---
title: "VA Data Quality Review Protocol / Report"
# author: 
date: "November 7, 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!--
This is a R Markdown comment:
You can use this notation to comment-out text. For example if you wnat this to be a pure protocol (i.e. just the protocol steps), you can comment-out the sections titles "Findings" below and they will not appear in your output document.
-->

## Group Work and ToT Follow-up

Over the last one and a half weeks in the "VA Data Quality Review and COD Assignment" Training of Trainers (5-16 November, 2018, Colombus, Ohio, USA) you have learned about the importance of VA raw data review, how to examine and review VA raw data using R, how the VA COD assignment algorithms work, how to run the VA COD algorithms, and how to interpret the outcome of the VA algorithm.

As Master Trainers for VA data quality review and COD assignment, you are now equipped to train others on these topics and you can apply the methods learned to the VA data collected in your country on a routine basis. For that application of VA data quality review steps (both pre- and post-algorithm), we ask you to develop a local protocol for routine application in your country and to work with local stakeholders to apply that protocol to your VA data on a routine (time interval will depend on the rate of VA data collection in your setting) basis.
To prepare you for the task of training others and the development of your local VA data quality review protocol, we have designed  this group exercise. You have been assigned to a group, to work together on the GroupWork.csv data set to apply the methods learned in Ohio and write a mock VA data quality review (pre- and post-algorithm) protocol and report (see **Findings** boxes below). In the course of the group exercise you will work together and teach each other to further solidify the learning from the Training of Trainers and you will develop the mock VA Data Quality Review Protocol / Report which you can use as a template for the protocol you will develop for application in your country.

We suggest you structure the VA Data Quality Review Protocol / Report into sections called VA Raw Data Examination and Cleaning, COD Assignment, and VA COD Examination. Please clearly number the steps of the protocol and comment all R code to allow for future use.

## VA Raw Data Examination and Cleaning

### Importing VA Data
Import the GroupWork.csv data set as a data frame called GWdata.
```{r data_import}

```

### Examining and Reviewing each Variable
```{r new variables, include=FALSE}
# This block will not be included in the VA Data Quality Review Protocol as echo=FALSE. You can use this block for R commands and data manipulations which you don't wnat to show in the Protocol but which you need to do in the "background".

```

#### Variable X 
```{r characterizing and checking variable X}

```

## Finding / Interpretation

Summarize the condition of each variable and what has to be done to it to make it useful.

### Variable: X, issue Y
```{r fixing-cleaning variable X issue Y}

```

### Variable: X, issue Z
```{r fixing-cleaning variable X issue Z}

```

## Running the VA Algorithm(s)
In the above steps the VA raw data was examined and cleaned and we now have a clean VA data set to run in the algorithm(s). 

*For countries using SmartVA Analyze, export the cleaned dataset to a csv using the following code (SmartVA-Analyze will not work with the GWdata; i.e. this part is only applicable for the countries using SmartVA-Analyze when using their own data to generate this Protocol / Report), run the algorithm and re-import the individual level output of SmartVA-Analyze (see file: cause-of-death) into R using the below code to then merge the input and output file to perform the VA COD Examination steps (see below).*
```{r exportRunImportSmart, include=FALSE}
##If you are suing SmartVA-Analzye, set include=TRUE and uncomment the below commnads.
#write.csv("FileLocation")
##Run the just exported CSV file in SmartVA-Analyze.
##Import the "cause-of-death.csv" file back into R.
#CODsmartVA<-read.csv("/cause-of-death.csv",header=TRUE,stringsAsFactors=FALSE)
```

## Run IntervA
Using the clean data set, we can now run the algorithm(s).
```{r InterVArun}
#openVA()
```

## Run InSilicoVA
```{r InSilicoVArun}
#openVA()
```

## VA COD Examination

Have a look at the VA COD assignments ...

### Merging Algorithm input and output data
```{r merge COD assignments with IDs and symptoms}

```

### Examining the merged and CODs and 

Explain how you will look at symptoms merged with CODs and assess plausibility.

#### Checking for unexpected CODs
```{r unexpected CODs}

```

#### Male versus Female causes
```{r male-female CODs}

```

### Plotting

Create various plots of the resulting COD assignments and CSMFs to support interpretation of the results

Plotting the CSMF from InterVA.
```{r interva plots}
#plot()
```

Plotting the CSMF from InSilicoVA.
```{r insilicova plots}
#plot()
```

Plotting the CSMF from all the algorithms together.
```{r csmf plots}
#plot()
```



